Anomalous waves triggered by abrupt depth changes: experimental measurements and deterministic-statistical truncated KdV theory

### 2019-10-04 paper2_JNLS/paper.tex
NM

A recent series of laboratory \cite{} and numerical investigations \cite{viotti2014} have demonstrated the emergence of anomalous wave statistics from abrupt variations in bottom topography. Since topographical variations are strictly one-dimensional, these studies can be viewed as offering a bare minimum set of conditions capable of generating anomalous wave statistics. In particular, the more complex mechanism of focusing by 2D bottom topography absent. The studies thus offer a paradigm system, with emergent anomalous statistical properties similar to that seen in more complex systems, but in a tractable context that is amenable to analysis.

Our particular focus is the laboratory experiments of Bolles et al.~\cite{}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In follow-up work, Majda and Qi \cite{majdaqi2019} analyzed more severely truncated systems of lower dimension that were shown to exhibit many of the same anomalous properties while enjoying a more tractable structure.

@@

% Figure: Experimental Diagram and Histograms
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.85 \linewidth]{Figs/ExpDiagStats.pdf}
\caption{
(a) Experimental schematic. 
(b)--(c) Surface displacement measured at a representative upstream and downstream location. (d)--(e) Corresponding histograms. Figure adapted from \cite{bolles2019}.}
\label{ExpDiagStats}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Data from column 5 in MasterTimeSeries, Delta theta = 1.38 degrees, which is in between runs 5 and 6.

	The free surface is illuminated by light-emitting diodes running along the bottom of the tank and is imaged from the sideview with a Nikon D3300 at 60 frames per second. The illumination technique, coupled with high pixel count of the camera, allows surface displacements to be resolved with accuracy better than 1/3 millimeter. Furthermore, these optical measurements permit extraction of wave statistics {\em continuously} in space, rather than at a few discrete locations. Further details of the experimental setup can be found in \cite{bolles2019}.

	Example measurements of free-surface displacements $\eta$ are shown in Figs.~\ref{ExpDiagStats}(b)--(c). These measurements are extracted from the images at two representative locations: one a short distance upstream of the ADC and one downstream of the ADC. Both signals exhibit a combination of periodic behavior and random fluctuations, with the dominant oscillations corresponding to the peak driving frequency of 2 Hz. The nature of the random fluctuations is revealed by the corresponding histograms shown in Fig.~\ref{ExpDiagStats}(d)--(e). The upstream measurements are symmetrically distributed about the mean, $\eta = 0$. In fact, Bolles et al.~found that these measurements closely follow a Gaussian distribution \cite{bolles2019}. The downstream measurements, however, are not symmetrically distributed and skew strongly towards positive displacement, $\eta > 0$. Bolles et al.~found that these measurements can be well described by a mean-zero gamma distribution \cite{bolles2019}. The slower decay rate of this downstream distribution indicates an elevated level of extreme surface displacement, i.e.~rogue waves. Bolles et al.~estimated that a rogue wave can be up to 65 times more likely in these experiments than if the displacements were Gaussian.

	The measurements shown in Fig.~\ref{ExpDiagStats} correspond to a paddle driving amplitude of $\Dphi = 1.38^{\circ}$, an upstream depth of $\dup = 12.5$ cm, a downstream depth of $\ddn = 3$ cm, giving a depth ratio of $\dratdn = \ddn/\dup= 0.24$. Bolles et al.~systematically varied the driving amplitude in the range $0.125^{\circ}$--$2^{\circ}$ to probe different regimes of wave behavior, from nearly linear waves to strongly nonlinear ones. They found that once a critical amplitude is exceeded (roughly $0.5^{\circ}$), the downstream statistics become anomalous and are robustly described by the gamma-distribution. No variation of the depth ratio was reported though. In the current manuscript, we also vary the downstream depth. We, however, will refer to the representative set of experiments from Fig.~\ref{ExpDiagStats} as a baseline for comparison.

\vsp{5}
TO DO: Describe the `representative' experiments in which $\dratdn = 0.24$ and $\epsup = 0.017$.

 % More details in \cite{}

 % Figure: Experimental Spatial Statistics
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.9 \linewidth]{Figs/ExpSpatialStats.pdf}
\caption{Spatial moments}
\label{ExpSpatialStats}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

With the experimental setup described and the theory outlined, we now present results comparing the two. Unless stated otherwise, all parameters used in the theory are taken directly from their experimental values listed in Table \ref{paramtable}. In particular, we set $\delup = 0.22$ in all tests, and for the `representative' experiments, we set $\dratdn = 0.24$ and $\epsup = 0.17$.

>>>>

A recent series of laboratory \cite{bolles2019} and numerical investigations \cite{viotti2014} have demonstrated the emergence of anomalous wave statistics from abrupt variations in bottom topography. Since topographical variations are strictly one-dimensional, these studies can be viewed as offering a bare minimum set of conditions capable of generating anomalous wave statistics. In particular, the more complex mechanism of focusing by 2D bottom topography absent. The studies thus offer a paradigm system, with emergent anomalous statistical properties similar to that seen in more complex systems, but in a tractable context that is amenable to analysis.

Our particular focus is the laboratory experiments of Bolles et al.~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, Majda et al.~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In follow-up work, Majda and Qi \cite{majdaqi2019} analyzed more severely truncated systems of lower dimension that were shown to exhibit many of the same anomalous properties while enjoying a more tractable structure.

@@

% Figure: Experimental Diagram and Histograms
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.85 \linewidth]{Figs/ExpDiagStats.pdf}
\caption{
(a) Experimental schematic. 
(b)--(c) Surface displacement measured at a representative upstream and downstream locations. (d)--(e) Corresponding histograms. Figure adapted from \cite{bolles2019}.}
\label{ExpDiagStats}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Data from column 5 in MasterTimeSeries, Delta theta = 1.38 degrees, which is in between runs 5 and 6.

	The free surface is illuminated by light-emitting diodes running along the bottom of the tank and is imaged from the sideview with a Nikon D3300 at 60 frames per second. The illumination technique, coupled with high pixel count of the camera, allows surface displacements to be resolved with accuracy better than 1/3 millimeter. Furthermore, these optical measurements permit extraction of wave statistics {\em continuously} in space, rather than at a few discrete locations, which is crucial for identifying regions of anomalous wave activity. Further details of the experimental setup can be found in \cite{bolles2019}.

	Example measurements of free-surface displacements $\eta$ are shown in Figs.~\ref{ExpDiagStats}(b)--(c). These measurements are extracted from the images at two representative locations: one a short distance (9 cm) upstream of the ADC and the other (15 cm) downstream of the ADC. Both signals exhibit a combination of periodic behavior and random fluctuations, with the dominant oscillations corresponding to the peak driving frequency of 2 Hz. The nature of the random fluctuations is revealed by the corresponding histograms shown in Fig.~\ref{ExpDiagStats}(d)--(e). The upstream measurements are symmetrically distributed about the mean, $\eta = 0$. In fact, Bolles et al.~found that these measurements follow a Gaussian distribution closely \cite{bolles2019}. The downstream measurements, however, are not symmetrically distributed and skew strongly towards positive displacement, $\eta > 0$. Bolles et al.~found that these measurements can be well described by a mean-zero gamma distribution \cite{bolles2019}. The slower decay rate of this downstream distribution indicates an elevated level of extreme surface displacement, i.e.~rogue waves. Bolles et al.~estimated that a rogue wave can be up to 65 times more likely in these experiments than if the displacements were Gaussian.

	The measurements shown in Fig.~\ref{ExpDiagStats} correspond to a paddle driving amplitude of $\Dphi = 1.38^{\circ}$, an upstream depth of $\dup = 12.5$ cm, and a downstream depth of $\ddn = 3$ cm, which gives the depth ratio $\dratdn = \ddn/\dup= 0.24$. Bolles et al.~systematically varied the driving amplitude in the range $0.125^{\circ}$--$2^{\circ}$ to probe different regimes of wave behavior, from nearly linear waves to strongly nonlinear ones. They found that once a critical amplitude is exceeded (roughly $\Dphi = 0.5^{\circ}$), the downstream statistics become anomalous and are robustly described by the gamma-distribution. Fig.~\ref{ExpSpatialStats} shows spatial statistics for several experiments of different driving amplitudes. The standard deviation of surface displacement $\etastd$ measures the characteristic amplitude of waves (Fig.~\ref{ExpSpatialStats}(a)). Notice that $\etastd$ increases with paddle amplitude, but does not significantly vary in space. In particular, $\etastd$ does not change significantly when waves cross the ADC. Skewness and (excess) kurtosis, however change markedly crossing the ADC once the critical driving amplitude is exceeded. We also show statistics of surface slope $\eta_x$.

 % Figure: Experimental Spatial Statistics
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[!ht]
\begin{center}
\includegraphics[width = 0.9 \linewidth]{Figs/ExpSpatialStats.pdf}
\caption{Wave statistics as they vary in space for several different experiments of different driving amplitudes (see legend). $\etastd$ does not vary significantly crossing the ADC (a), but skewness (b) and kurtosis (c) show a strong response to the ADC. }
\label{ExpSpatialStats}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

	Bolles et al.~systematically explored the effect of varying the driving amplitude on the anomalous statistics for a fixed depth ratio. Here, we further explore the effect of driving amplitude, especially in conjunction with theoretical developments, and we will also report experiments in which the depth ratio was varied. We will refer to the experiments displayed in Fig.~\ref{ExpDiagStats} as the reference experiments, so that we have a baseline for comparison. In these reference experiments, the driving amplitude is $\Dphi = 1.38^{\circ}$ (intermediate between the cases $\Dphi = 1.25^{\circ}$ and $1.5^{\circ}$ shown in Fig.~\ref{ExpDiagStats}(a)), the depth ratio is $\dratdn = 0.24$, and the characteristic wave amplitude is $\etastd = 0.21$ cm.

%$\epsup = 0.017$.

@@

With the experimental setup described and the theory outlined, we now present results comparing the two. Unless stated otherwise, all parameters used in the theory are taken directly from their experimental values listed in Table \ref{paramtable}. In particular, we set $\delup = 0.22$ in all tests, and for the `representative' experiments, we set $\dratdn = 0.24$ and $\epsup = 0.017$.
